---
title: "RGEOde Tutorial: Principal Component Analysis (PCA)"
subtitle: "Example using Motor Trend Car Data"
output: html_document
---

# Data source:
The data was extracted from the 1974 Motor Trend US magazine, and comprises fuel consumption and **10 variables** related to automobile design and performance for **32 cars** (1973–74 models).

# 10 Variables:   
 - mpg	= Miles/gallon  
 - cyl	= Number of cylinders  
 - disp = Displacement (cubic inches)  
 - hp = Gross horsepower  
 - drat = Rear axle ratio  
 - wt = Weight (1000 lbs)  
 - qsec = 1/4 mile time  
 - vs = V/S  
 - am = Transmission (0 = automatic, 1 = manual)  
 - gear = Number of forward gears  
 - carb = Number of carburetors

# Setup

### Date and time that analysis started
```{r setup_time}
Sys.time()
```

## Install R packages (only do once)  
*If they are not already downloaded.*  
```{r setup_pkgs_install, eval=FALSE}
### R packages at CRAN
install.packages(c("ggplot2", "knitr"))
```

## Load R packages  
*These packages provide us with the functions we will use in our analysis.*   
```{r setup_pkgs_load}
library(ggplot2)
library(knitr)
```

## RMarkdown report options
```{r setup_opts}
### Do not print R code in report
opts_chunk$set(echo = FALSE, comment=NA)

### Do not print warnings and messages in report
opts_chunk$set(warning = FALSE, message = FALSE)

### Keep track of RMarkdown analysis progress
opts_knit$set(progress = TRUE, verbose = TRUE)
```

## Load custom pca plotting function 
```{r setup_fxn}
### PCA plotting function (for prcomp) modified from:
### ggbiplot R package (download from GitHub website using devtools)
#install.packages("devtools")
#library("devtools")
#install_github("vqv/ggbiplot")

pcaplot <- function (pcobj, choices = 1:2, scale = 1, 
    obs.scale = 1 - scale, var.scale = scale, circle.prob = 0.68,
    groups = NULL, ellipse = TRUE, ellipse.prob = 0.68, 
    labels = NULL, labels.size = 3, alpha = 1, 
    var.axes = FALSE, var.color = "darkblue", 
    varname.size = 3, varname.adjust = 1.5, varname.abbrev = FALSE, ...) 
{
  library(ggplot2)
  library(plyr)
  library(scales)
  library(grid)
  stopifnot(length(choices) == 2)
    
  nobs.factor <- sqrt(nrow(pcobj$x) - 1)
  d <- pcobj$sdev
  u <- sweep(pcobj$x, 2, 1/(d * nobs.factor), FUN = "*")
  v <- pcobj$rotation

  choices <- pmin(choices, ncol(u))
  df.u <- as.data.frame(sweep(u[, choices], 2, d[choices]^obs.scale, FUN = "*"))
  v <- sweep(v, 2, d^var.scale, FUN = "*")
  df.v <- as.data.frame(v[, choices])
  names(df.u) <- c("xvar", "yvar")
  names(df.v) <- names(df.u)
  
  df.u <- df.u * nobs.factor
  r <- sqrt(qchisq(circle.prob, df = 2)) * prod(colMeans(df.u^2))^(1/4)
  v.scale <- rowSums(v^2)
  df.v <- r * df.v/sqrt(max(v.scale))
  
  if (obs.scale == 0) u.axis.labs <- paste("PC", choices, sep = "")
  else u.axis.labs <- paste("PC", choices, sep = "")
  
  u.axis.labs <- paste(
    u.axis.labs, 
    sprintf("(%0.1f%% variance)", 100 * pcobj$sdev[choices]^2/sum(pcobj$sdev^2)))
  if (!is.null(labels)) df.u$labels <- labels
  if (!is.null(groups)) df.u$groups <- groups
  
  if (varname.abbrev) df.v$varname <- abbreviate(rownames(v))
  else df.v$varname <- rownames(v)
  df.v$angle <- with(df.v, (180/pi) * atan(yvar/xvar))
  df.v$hjust <- with(df.v, (1 - varname.adjust * sign(xvar))/2)
  
  g <- ggplot(data = df.u, aes(x = xvar, y = yvar)) + 
    labs(x = u.axis.labs[1], y = u.axis.labs[2]) + 
    theme_bw()
  
  if (var.axes) {
    g <- g + geom_segment(
      data = df.v, 
      aes(x = 0, y = 0, xend = xvar, yend = yvar),
      arrow = arrow(length = unit(1/2, "picas")), 
      color = muted(var.color))
  }
  
  if (!is.null(df.u$labels)) {
    if (!is.null(df.u$groups)) g <- g + geom_text(aes(label = labels, color = groups), size = labels.size)
    else g <- g + geom_text(aes(label = labels), size = labels.size)
  } else {
    
    if (!is.null(df.u$groups)) g <- g + geom_point(aes(color = groups), alpha = alpha)
    else g <- g + geom_point(alpha = alpha)
  }
  
  if (!is.null(df.u$groups) && ellipse) {
    theta <- c(seq(-pi, pi, length = 50), seq(pi, -pi, length = 50))
    circle <- cbind(cos(theta), sin(theta))
    ell <- ddply(df.u, "groups", function(x) {
      if (nrow(x) <= 2) return(NULL)
      sigma <- var(cbind(x$xvar, x$yvar))
      mu <- c(mean(x$xvar), mean(x$yvar))
      ed <- sqrt(qchisq(ellipse.prob, df = 2))
      data.frame(sweep(circle %*% chol(sigma) * ed, 2, mu, FUN = "+"), groups = x$groups[1])
      })
    names(ell)[1:2] <- c("xvar", "yvar")
    g <- g + geom_path(data = ell, aes(color = groups, group = groups))
  }
  
  if (var.axes) {
    g <- g + geom_text(
      data = df.v, 
      aes(label = varname, x = xvar, y = yvar, angle = angle, hjust = hjust), 
      color = var.color, 
      size = varname.size)
  }
  return(g)
}
```

# Load the dataset  
It is included with R  
```{r data_load}
data(mtcars)
```

# Table 1. Print the dataset
```{r data_print}
print(mtcars)
```

# Run principal component analysis (PCA) on the dataset    
```{r pca_run}
mypca <- prcomp(mtcars, 
                scale. = TRUE, # scale variables to have unit variance (generally advised)
                center = TRUE) # shift variables to be zero-centered
```

# Table 2. PCA results  
Print standard deviations, variance explained, and cumulative variance explained
```{r pca_summary}
summary(mypca)
```

# Figure 1: PCA results  
For each car, plot values where x = PC1 and y = PC2
```{r pca_plot}
library("ggbiplot")
plot1 <- pcaplot(mypca)
print(plot1)
```

# Figure 2: Outliers  
Add car names to plot, to identify outlier cars
```{r pca_plot_name_outliers}
plot2 <- pcaplot(mypca, labels = rownames(mtcars)) +
  ggtitle("Figure 2. Outliers") # add title
print(plot2)
```

# Figure 3: Measures (variables) of automobile design and performance  
Add measure information to plot, to help explain outlier cars   
```{r pca_plot_name_outliers_variables}
plot3 <- pcaplot(mypca, labels = rownames(mtcars), var.axes = TRUE, var.color = "darkblue") +
  ggtitle("Figure 3. Outliers & Variables") # add title
print(plot3)

### Save (optional)
ggsave("pca_plot_outliers_variables.png", width = 6, height = 4, dpi = 300)
```

## Methods   
```{r methods}
### Date and time that analysis finished
Sys.time()

### How long the analysis took
proc.time()

### Packages used (and their versions) for reproducibility
sessioninfo::session_info()
```
